Analysing the Performance of Shallow-Water Models with the Roofline Model

### 2016-01-22 docs/papers/roofline_model/gocean_roofline.tex
roofline: added stream2 results for Archer

Interpreting the results of the STREAM benchmark in order to get
bandwidth figures for all levels of the cache hierarchy is not
straightforward. We therefore used the STREAM2 benchmark which is
intended for this purpose. In Figure~\ref{FIG_stream2_desktop} we plot
the results obtained from the DAXPY component of the STREAM2 benchmark
on the E5-1620 CPU. The plateaux in bandwidth for L2, L3 and main
memory are clear but that for L1 is considerably noisier. The final
values we have used in the roofline plots are given in
Table~\ref{TAB_stream_linpack}.

\begin{figure}
\includegraphics{stream2_desktop}
\caption{Results from running the STREAM2 benchmark on the E5-1620 CPU.}
\label{FIG_stream2_desktop}
\end{figure}

\begin{table}
\begin{tabular}{|l|c|c|c|c|}
\hline
  \multicolumn{5}{|c|}{E5-1620 v.~2, 3.7GHz} \\
  \hline
LINPACK (GFLOPS) & \multicolumn{4}{c|}{28.32} \\
\hline
 &  L1 Cache & L2 Cache & L3 Cache & Main memory \\
 &   32 KB   & 256 KB   & 10 MB    &  --         \\
\hline
STREAM2 (GB/s)   &  117  &  65.8  & 46.7 &  20.5 \\
  \hline
\multicolumn{5}{|c|}{E5-2697 v.~2, 2.7GHz}  \\
\hline
LINPACK (GFLOPS) & \multicolumn{4}{c|}{25.41} \\
\hline
 & L1 Cache & L2 Cache & L3 Cache & Main memory \\
 &   32 KB  & 256 KB   & 30 MB    &  --         \\
\hline
STREAM2 (GB/s) &   &   &  &  \\
\hline
\end{tabular}
\caption{The STREAM and LINPACK benchmark results for a single thread
  running on each of the two CPUs considered here.}
\label{TAB_stream_linpack}
\end{table}

@@

The likwid tool (\url{https://github.com/RRZE-HPC/likwid}) gives
access to Hardware Performance Counters (HWPCs) on Intel and AMD CPUs.
However, there are issues with the accuracy of these counters,
especially for FLOP-related measurements on Ivybridge architectures. A
tool is included in the likwid distribution that tests the accuracy of
HWPC-derived data by running micro-benchmarks for which the FLOP count
and memory traffic are known:
\url{https://code.google.com/p/likwid/wiki/IvyBridgeVerification}.
The results of doing this on an Intel Xeon E5-1620 v2 CPU are shown in
Figures~\ref{FIG_flops_dp_stream_test}
and~\ref{FIG_flops_dp_triad_test}. It appears that once the memory
footprint spills out of L1 cache (which all of our nemolite2d
benchmark configurations do), the performance counter results
significantly overestimate the FLOPs performed.

\begin{figure}
\centering
\includegraphics[width=120mm]{FLOPS_DP_stream_ivybridgeEP.pdf}
\caption{Results of the likwid accuracy test for the FLOPS\_DP group
  and the stream benchmark on an Intel Xeon E5-1620 v2.}
\label{FIG_flops_dp_stream_test}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=120mm]{FLOPS_DP_triad_ivybridgeEP.pdf}
\caption{Results of the likwid accuracy test for the FLOPS\_DP group
  and the triad benchmark on an Intel Xeon E5-1620 v2.}
\label{FIG_flops_dp_triad_test}
\end{figure}

@@

We can also count the number of floating-point arithmetic operations
in this code fragment which gives us 12 multiplications and 12
additions. $AI$ for this code fragment is then $24*7/640 = 0.26$
FLOPs/byte.  If we do the same analysis for the (longer) u-Momentum
kernel of nemolite2d then we find that there are 51 addition
operations, 56 multiplication/division operations and one call to
sin(). There are 31 distinct memory references. This gives an $AI$ of
approximately $(107+13)*7/(31*64)$ or 0.42 FLOPs/byte.

\begin{table}
  \begin{tabular}{l|c|c|c}
                           & Shallow & \multicolumn{2}{c}{nemolite2d} \\
                           & 1st Loop nest & Continuity & u-Momentum \\
\hline                                
FLOPs                      & 24      &    14      &   107+13   \\
Distinct memory refs       & 10      &    10      &   31       \\
$AI$ (FLOPs/byte)          & 0.26    &  0.153     &   0.42     \\
Working set size (bytes)   & $56n^2$ &  $64n^2$   &            \\
\end{tabular}
\caption{Details of the two nemolite2d kernels investigated with the
  roofline model. $n$ is the problem size -- the length of one side of
  the square domain.}
\label{TAB_kernel_details}
\end{table}

@@

It feels as though there is a need for a tool that addresses some of
these issues.

>>>>

Interpreting the results of the STREAM benchmark in order to get
bandwidth figures for all levels of the cache hierarchy is not
straightforward. We therefore used the STREAM2 benchmark which is
intended for this purpose. In Figure~\ref{FIG_stream2} we plot
the results obtained from the DAXPY component of the STREAM2 benchmark
on the E5-1620 CPU. The plateaux in bandwidth for L2, L3 and main
memory are clear but that for L1 is considerably noisier. The final
values we have used in the roofline plots are given in
Table~\ref{TAB_stream_linpack}.

\begin{figure}
\includegraphics{stream2_desktop}
\includegraphics{stream2_archer}
\caption{Results from running the STREAM2 benchmark on the E5-1620
  (top) and E5-2697 (bottom) CPUs.}
\label{FIG_stream2}
\end{figure}

\begin{table}
\begin{tabular}{|l|c|c|c|c|}
\hline
  \multicolumn{5}{|c|}{E5-1620 v.~2, 3.7GHz} \\
  \hline
LINPACK (GFLOPS) & \multicolumn{4}{c|}{28.32} \\
\hline
 &  L1 Cache & L2 Cache & L3 Cache & Main memory \\
 &   32 KB   & 256 KB   & 10 MB    &  --         \\
\hline
STREAM2 (GB/s)   &  117  &  65.8  & 46.7 &  20.5 \\
  \hline
\multicolumn{5}{|c|}{E5-2697 v.~2, 2.7GHz}  \\
\hline
LINPACK (GFLOPS) & \multicolumn{4}{c|}{25.41} \\
\hline
 & L1 Cache & L2 Cache & L3 Cache & Main memory \\
 &   32 KB  & 256 KB   & 30 MB    &  --         \\
\hline
STREAM2 (GB/s) & 108 & 54.0 & 37.7  &  17.6 \\
\hline
\end{tabular}
\caption{The STREAM and LINPACK benchmark results for a single thread
  running on each of the two CPUs considered here.}
\label{TAB_stream_linpack}
\end{table}

@@

The likwid tool (\url{https://github.com/RRZE-HPC/likwid}) gives
access to Hardware Performance Counters (HWPCs) on Intel and AMD CPUs.
However, there are issues with the accuracy of these counters,
especially for FLOP-related measurements on Ivybridge architectures. A
tool is included in the likwid distribution that tests the accuracy of
HWPC-derived data by running micro-benchmarks for which the FLOP count
and memory traffic are known:
\url{https://code.google.com/p/likwid/wiki/IvyBridgeVerification}.
The results of doing this on an Intel Xeon E5-1620 v2 CPU are shown in
Figure~\ref{FIG_flops_dp_test}. It appears that once the memory
footprint spills out of L1 cache (which all of our nemolite2d
benchmark configurations do), the performance counter results
significantly overestimate the FLOPs performed.

\begin{figure}
\centering
\includegraphics[width=120mm]{FLOPS_DP_stream_ivybridgeEP.pdf}
\includegraphics[width=120mm]{FLOPS_DP_triad_ivybridgeEP.pdf}
\caption{Results of the likwid accuracy test for the FLOPS\_DP group
  and the stream (top) and triad (bottom) benchmark on an Intel Xeon
  E5-1620 v2.}
\label{FIG_flops_dp_test}
\end{figure}

@@

We can also count the number of floating-point arithmetic operations
(FLOPs) in this code fragment which gives us 12 multiplications and 12
additions. The $AI$ for this code fragment is then $24*7/640 = 0.26$
FLOPs/byte.  If we do the same analysis for the (longer) u-Momentum
kernel of nemolite2d then we find that there are 51 addition
operations, 56 multiplication/division operations and one call to
sin(). There are 31 distinct memory references. This gives an $AI$ of
approximately $(107+13)*7/(31*64)$ or 0.42 FLOPs/byte (where we assume
the call to {\tt sin()} costs 13 FLOPs). The results of this analysis
are summarised in Table~\ref{TAB_kernel_details}.

\begin{table}
\begin{tabular}{|l|c|c|c|}
\hline
Code                       & Shallow & \multicolumn{2}{c|}{nemolite2d} \\
\hline
Kernel                     & 1st loop nest & Continuity & u-Momentum \\
\hline                                
FLOPs                      & 24      &    14      &   107+13   \\
Distinct memory refs       & 10      &    10      &   31       \\
$AI$ (FLOPs/byte)          & 0.26    &  0.153     &   0.42     \\
Working set size (bytes)   & $56n^2$ &  $64n^2$   &   $136n^2$ \\
\hline
\end{tabular}
\caption{Details of the kernels investigated with the roofline
  model. $n$ is the problem size -- the length of one side of the
  square domain.}
\label{TAB_kernel_details}
\end{table}

@@

It feels as though there is a need for a tool that addresses some of
these issues. It could be that Intel's Vectorization Advisor will be
useful and this requires investigation.
